\documentclass[noanswers]{exam}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}

\title{Problem Set 3\\Topology and Differentiation}
\author{Paulo Fagandini}
\date{August 2018}

\begin{document}

\maketitle

\begin{questions}

\section*{Topology and Continuity}

\question  Show that the open ball around $x_0\in\mathbb{R}$, $B(x_0,r)$ with $r>0$ is an open set.

\question Let $A\subseteq\mathbb{R}^n$. Consider the collection of sets $A_i\subseteq\mathbb{R}^n$, $i\in I$. Show that

\begin{parts}
\part $int(A)\subseteq A$
\part $A$ is open, if and only if $int(A)=A$.
\part If $A_i$ is open, for any $i$, then $\bigcup_{i\in I} A_i$ is open.
\part If $A_i$ is open, for any $i$ and $I$ is finite, then $\bigcap_{i\in I} A_i$ is open.
\part $\phi$ and $\mathbb{R}^n$ are open sets and closed at the same time.
\part If $A_i$ is closed for any $i$, then $\cap_{i\in I} A_i$ is closed.
\part If $A_i$ is closed for every $i$, and $I$ is finite, then $\cup_{i=I} A_i$ is closed.
\end{parts}

\question Let $A_i=\left(\frac{-1}{i},\frac{1}{i}\right)$. show that $\cup_{i\in\mathbb{N}}A_i=\{0\}$, and show that $\{0\}$ is not open.

\question Show that the sequence $x_t=K$, with $K\in\mathbb{R}$ constant, then $x_t$ converges, and its limit is $K$.

\question Show that if the sequence $x_t$ converges, then $X=\{x_t|t\in\mathbb{N}\} $ is bounded.

\question Show that if $x_t\rightarrow x_0$ and $y_t\rightarrow x_0$, both sequences in $\mathbb{R}$, then $(x_t+y_t)\rightarrow(x_0+y_0)$.

\question Show that if the sequence $x_t\rightarrow x^*$ is such that, for any $t$ $x_t>0$, then $x^*\geq 0$.

\question Show that if $K$ is compact, and $A$ is closed, then $K\cap A$ is compact.

\question Show that if $a_n\rightarrow a$, then $A=\{a_n, n\in\mathbb{N}\}$ is bounded. Discuss if $A$ is closed or not.

\question Let $x_t$ a sequence such that $\left|\frac{x_{t+1}}{x_t}\right|\rightarrow L$. Show that if $0<L<1$ then the sequence $x_t$ converges.

\question Let $x_t$ be a sequence such that $\sqrt[t]{|x_t|}\rightarrow L$. Show that if $0<L<1$ then $x_t$ converges.

\question Show that if $f:\mathbb{R}\rightarrow\mathbb{R}$ is continuous, and $f(x_0)>0$, then there is $r>0$, such that for $x\in B(x_0,r)$ open, then $f(x)>0$.

\question Let $f:\mathbb{R}\rightarrow\mathbb{R}$ continuous, let $B\subseteq\mathbb{R}$ open, show that $f^{-1}(B)$ is also open.

\section*{Differentiation}

\question Show that $f(x)=x^2$ is differentiable. Find its derivative.

\question Show that $f(x)=|x|$ is not differentiable.

\question Find regions where the function $f(x)=x^2-3x^2+x$ is increasing.

\question Find the Taylor series, of order $k$ and around $x_0=0$, of the following functions.

\begin{parts}
\part $f(x)=e^x$
\part $f(x)=\sin(x)$
\part $f(x)=\cos(x)$
\end{parts}

\question The Taylor series for a function $f:\mathbb{R}\rightarrow\mathbb{R}$ of order 2, and around $x_0$ corresponds to:

\begin{align*}
    T_2(x_0,f)(x)=f(x_0)+\nabla^t f(x_0)(x-x_0)+\frac{1}{2}(x-x_0)^tH(f,x_0)(x-x_0)
\end{align*}

Find $T_2(x_0,f)$ when $x_0=(1,1)$ and $f(x_1,x_2)=x_1^\alpha x_2^\beta$.

\question Using L'H\^ opital's rule compute the following limits:

\begin{parts}

\part $\lim_{x\rightarrow 2}\frac{e^{x^2}-e^4}{x-2}$
\part $\lim_{x\rightarrow\infty}\frac{\sqrt[3]{x}}{\log(x)}$

\part $\lim_{x\rightarrow a}\frac{x-a}{\ln(x)-\ln(a)}$

\part $\lim_{x\rightarrow 0}\frac{e^x-1}{x}$

\end{parts}

\question Show that $f:\mathbb{R}\rightarrow\mathbb{R}$, where $f(x)=\frac{1}{2}x$ has a unique fixed point.

\question Does the function $f:[a,b]\rightarrow[a,b]$, with $a$ and $b$ finite, such that $f(x)=\frac{h(g(x)+1)}{\phi(x)}$, with $h,g$, and $\phi$ continuous, has a fixed point? If so, is this unique?

\end{questions}


\end{document}
