---
title: "Continuous Probabilistic Models"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School – Polytechnic University of Lisbon"
format: 
  revealjs:
    theme: default
    logo: images/iscal_logo.svg
    slide-number: true
    cross-ref: true
    incremental: true
    math: 
      method: mathjax
    transition: slide
    footer: "Statistics I"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.75em
    progress: true
editor: source
execute:
  enabled: true
engine: knitr
---

## Disclaimer

These slides are a free translation and adaptation from the slide deck for Estatística I by Prof. Sandra Custódio and Prof. Teresa Ferreira from the Lisbon Accounting and Business School - Polytechnical University of Lisbon.

# Uniform Distribution

## Continuous Uniform Distribution

A r.v. $X$ follows a uniform distribution in $[a,b]\subset\mathbb{R}$ with $-\infty<a<b<\infty$, if its probability density is given by:

$$
f_X(x)=\begin{cases}
\frac{1}{b-a} & a\leq x \leq b\\
0 & otherwise
\end{cases}
$$

We write $X\sim U[a,b]$

## Continuous Uniform Distribution

```{r, engine='tikz'}
#| fig-align: center

\begin{tikzpicture}

\draw[<->] (-2,0) -- (4,0)node[below]{$x$};
\draw[->] (0,0) -- (0,4)node[left]{$f_X$};

\draw (0,1) node[left]{$\frac{1}{b-a}$};

\draw[red,thick] (-1.75,0) -- (1,0)node[below]{$a$};
\draw[red,thick] (3,0)node[below]{$b$} -- (3.75,0);
\draw[red,thick] (1,1) -- (3,1);

\end{tikzpicture}

```

## Continuous Uniform Distribution

The cumulative distribution function is given by:

$$
F_X(x)=\begin{cases}
0 & x< a\\
\frac{x-a}{b-a} & a\leq x < b\\
1 & x\geq b
\end{cases}
$$

## Continuous Uniform Distribution

Note, this distribution is *symmetric*, and its first two moments are:

* $E[X]=\frac{a+b}{2}$
* $V[X]=\frac{(b-a)^2}{12}$

## Continuous Uniform Distribution - Example {#unif-example-1}

The length of small spots in a TV network is a r.v. $X$ distributed $U[5,12]$.

1. Find its distribution function.
2. What is the probability of a small spot last at least 7 seconds?
3. The probability that a small spot lasts more than 6 seconds, given it never lasts more than 10 seconds is?
4. Find $E[X]$ and $V[X]$

[Solution](#unif-example-sol-1)

## Continuous Uniform Distribution - Example {#unif-example-2}

Let the r.v. $X$ be distributed $U[2,b]$ with $b>2$. What value must $b$ take to make $P(3\leq X\leq 5)=0.4$?

[Solution](#unif-example-sol-2)

# Exponential Distribution

## Exponential Distribution

The exponential distribution is rooted in the Poisson distribution, reflecting the waiting time between events originated according to a Poisson process.

Nevertheless, we can apply the exponential distribution to many other phenomena.

## Exponential Distribution

A r.v $X$ is distributed exponentially, with parameter $\lambda>0$, $X\sim Exp(\lambda)$, if its probability density function is given by:

$$
f_X(x)=\begin{cases}
\lambda e^{-\lambda x} & x\geq 0\\
0 & x< 0
\end{cases}
$$

> $\lambda$ can be interpreted as the expected waiting time (or space) between events.

## Exponential Distribution

```{r, engine='tikz'}
#| fig-align: center

\begin{tikzpicture}

\draw[<->] (-2,0) -- (4,0)node[below]{$x$};
\draw[->] (0,0) -- (0,4)node[left]{$f_X$};

\draw[red,thick,domain=0.001:3,samples=100] plot(\x, {2*exp(-2*\x)});
\draw[red,thick] (-1,0)--(0,0);

\end{tikzpicture}

```

## Exponential Distribution

The cumulative probability function is:

$$
F_X(x)=\begin{cases}
0 & x< 0\\
1-e^{-\lambda x} & x\geq 0
\end{cases}
$$

And its moments are:

* $E[X]=\frac{1}{\lambda}$
* $V[X]=\frac{1}{\lambda^2}$

# Normal Distribution

# Appendix

## Uniform Exercise {#unif-example-1-sol}

1. $$
F_X(x)=\begin{cases}
0 & x<5\\
\frac{x-5}{12-5} & 5\leq x < 12\\
1 & x\geq 12 
\end{cases}
2. $F_X(12)-F_X(7)=1-\frac{2}{7}=\frac{5}{7}$
3. $\frac{F_X(10)-F_X(6)}{F_X(10)}=\frac{5/7-1/7}{5/7}=4/5$
4. $E[X]=\frac{5+12}{2}=8.5$, $V[X]=\frac{(12-5)^2}{12}=\frac{49}{12}\approx 4 \Rightarrow \sigma_X\approx\sqrt{4}\approx 2$

[Back to the exercise](#unif-example-1)

## Uniform Exercise {#unif-example-2-sol}

1. $P(3\leq X\leq 5)=0.4\ \Rightarrow\ F_X(5)-F_X(3)=0.4$
2. $\frac{5-2}{b-2}-\frac{3-2}{b-2}=0.4\ \Rightarrow\ \frac{3}{b-2}-\frac{1}{b-2}=0.4$
4. $\frac{2}{b-2}=0.4\ \Rightarrow\ \frac{1}{b-2}=0.2$, but $0.2 = \frac{2}{10} = \frac{1}{5}$
5. $\frac{1}{b-2}=\frac{1}{5}\ \Rightarrow\ 5=b-2\ \Rightarrow\ b=7$



[Back to the exercise](#unif-example-2)