---
title: "Continuous Probabilistic Models"
author: "Paulo Fagandini"
institute: "Lisbon Accounting and Business School – Polytechnic University of Lisbon"
format: 
  revealjs:
    theme: default
    logo: images/iscal_logo.svg
    slide-number: true
    cross-ref: true
    incremental: true
    math: 
      method: mathjax
    transition: slide
    footer: "Statistics I"
    from: markdown+emoji
    code-overflow: wrap
    code-block-font-size: 0.75em
    progress: true
editor: source
execute:
  enabled: true
engine: knitr
---

## Disclaimer

These slides are a free translation and adaptation from the slide deck for Estatística I by Prof. Sandra Custódio and Prof. Teresa Ferreira from the Lisbon Accounting and Business School - Polytechnical University of Lisbon.

# Uniform Distribution

## Continuous Uniform Distribution

A r.v. $X$ follows a uniform distribution in $[a,b]\subset\mathbb{R}$ with $-\infty<a<b<\infty$, if its probability density is given by:

$$
f_X(x)=\begin{cases}
\frac{1}{b-a} & a\leq x \leq b\\
0 & otherwise
\end{cases}
$$

We write $X\sim U[a,b]$

## Continuous Uniform Distribution

```{r, engine='tikz'}
#| fig-align: center

\begin{tikzpicture}

\draw[<->] (-2,0) -- (4,0)node[below]{$x$};
\draw[->] (0,0) -- (0,4)node[left]{$f_X$};

\draw (0,1) node[left]{$\frac{1}{b-a}$};

\draw[red,thick] (-1.75,0) -- (1,0)node[below]{$a$};
\draw[red,thick] (3,0)node[below]{$b$} -- (3.75,0);
\draw[red,thick] (1,1) -- (3,1);

\end{tikzpicture}

```

## Continuous Uniform Distribution

The cumulative distribution function is given by:

$$
F_X(x)=\begin{cases}
0 & x< a\\
\frac{x-a}{b-a} & a\leq x < b\\
1 & x\geq b
\end{cases}
$$

## Continuous Uniform Distribution

Note, this distribution is *symmetric*, and its first two moments are:

* $E[X]=\frac{a+b}{2}$
* $V[X]=\frac{(b-a)^2}{12}$

## Continuous Uniform Distribution - Example {#unif-example-1}

The length of small spots in a TV network is a r.v. $X$ distributed $U[5,12]$.

1. Find its distribution function.
2. What is the probability of a small spot last at least 7 seconds?
3. The probability that a small spot lasts more than 6 seconds, given it never lasts more than 10 seconds is?
4. Find $E[X]$ and $V[X]$

[Solution](#unif-example-sol-1)

## Continuous Uniform Distribution - Example {#unif-example-2}

Let the r.v. $X$ be distributed $U[2,b]$ with $b>2$. What value must $b$ take to make $P(3\leq X\leq 5)=0.4$?

[Solution](#unif-example-sol-2)

# Exponential Distribution

## Exponential Distribution

The exponential distribution is rooted in the Poisson distribution, reflecting the waiting time between events originated according to a Poisson process.

Nevertheless, we can apply the exponential distribution to many other phenomena.

## Exponential Distribution

A r.v $X$ is distributed exponentially, with parameter $\lambda>0$, $X\sim Exp(\lambda)$, if its probability density function is given by:

$$
f_X(x)=\begin{cases}
\lambda e^{-\lambda x} & x\geq 0\\
0 & x< 0
\end{cases}
$$

> $\lambda$ can be interpreted as the expected waiting time (or space) between events.

## Exponential Distribution

```{r, engine='tikz'}
#| fig-align: center

\begin{tikzpicture}

\draw[<->] (-2,0) -- (4,0)node[below]{$x$};
\draw[->] (0,0) -- (0,4)node[left]{$f_X$};

\draw[red,thick,domain=0.001:3,samples=100] plot(\x, {2*exp(-2*\x)});
\draw[red,thick] (-1,0)--(0,0);

\end{tikzpicture}

```

## Exponential Distribution

The cumulative probability function is:

$$
F_X(x)=\begin{cases}
0 & x< 0\\
1-e^{-\lambda x} & x\geq 0
\end{cases}
$$

And its moments are:

* $E[X]=\frac{1}{\lambda}$
* $V[X]=\frac{1}{\lambda^2}$

## Exponential Distribution

::: {.callout}

## Property

Lack of memory of the exponential distribution:

Let the r.v. $X\sim Exp(\lambda)$, then:

$$P(X>x+h|X>x)=P(X>h)$$

with $x,h>0$

:::

Considering the survival application of this distribution, this property states that, the time left to leave is independent of what it already lived.

## Exponential Distribution - Example {#exp-example-1}

In a factory, the execution time of a piece is random variable distributed exponentially with expected value of 5 minutes.

1. For this distribution, what is the value of $\lambda$?
2. You know that the piece has been in execution by at least 2 minutes. What is the probability that 4 more minutes are necessary to complete the piece?
3. Picking randomly 5 pieces. What is the probability that 2 out of the 5 were produced in less than 4 minutes?

[Solution](#exp-example-sol-1)

## Exponential Distribution - Example {#exp-example-2}

The time it takes until the first consultation, and between consultations, in the clinic of Dr. Shawn are independent and distributed exponentially with $\lambda=0.1$.

What is the probability that no consultation occurs before the first 10 minutes?

[Solution](#exp-example-sol-2)

# Normal Distribution

## References

::: {.nonincremental}

* Murteira, B.; Silva Ribeiro, C.; Andrade e Silva, J. & Pimenta, C., Introdução à Estatística, Escolar Editora, McGraw-Hill, 2010
* Paulino, C. D. & Branco, J. A. (2005). Exercícios de Probabilidade e Estatística. Escolar Editora
* Pimenta, F., Andrade e Silva, J.; Silva Ribeiro, C. & Murteira, B., Introdução à Estatística – 3ª Edição, Escolar Editora, 2015
* Ferreira, T., Custódio, S.G., Modelos Probabilísticos – Síntese Teórica e Exercícios Resolvidos, Edições Sílabo (1ª Edição), 2023 

:::

# Appendix

## Uniform Exercise {#unif-example-1-sol}

::: {.nonincremental}

1. 
$$
F_X(x)=
\begin{cases}
0 & x<5\\
\frac{x-5}{12-5} & 5\leq x < 12\\
1 & x\geq 12 
\end{cases}
$$
2. $F_X(12)-F_X(7)=1-\frac{2}{7}=\frac{5}{7}$
3. $\frac{F_X(10)-F_X(6)}{F_X(10)}=\frac{5/7-1/7}{5/7}=4/5$
4. $E[X]=\frac{5+12}{2}=8.5$, $V[X]=\frac{(12-5)^2}{12}=\frac{49}{12}\approx 4 \Rightarrow \sigma_X\approx\sqrt{4}\approx 2$

:::

[Back to the exercise](#unif-example-1)

## Uniform Exercise {#unif-example-2-sol}

::: {.nonincremental}

1. $P(3\leq X\leq 5)=0.4\ \Rightarrow\ F_X(5)-F_X(3)=0.4$
2. $\frac{5-2}{b-2}-\frac{3-2}{b-2}=0.4\ \Rightarrow\ \frac{3}{b-2}-\frac{1}{b-2}=0.4$
4. $\frac{2}{b-2}=0.4\ \Rightarrow\ \frac{1}{b-2}=0.2$, but $0.2 = \frac{2}{10} = \frac{1}{5}$
5. $\frac{1}{b-2}=\frac{1}{5}\ \Rightarrow\ 5=b-2\ \Rightarrow\ b=7$

:::

[Back to the exercise](#unif-example-2)

## Exponential Exercise {#exp-example-sol-1}

::: {.nonincremental}

1. $\lambda=\frac{1}{5}=0.2$. On average 2 pieces are produced every minute. $X\sim Exp(0.2)$
2. $P(X>6|X>2)=P(X>4)=1-P(T<4)=$ $1-\left(1-e^{-0.2\times 4}\right)=e^{-0.8}\approx 0.4493$
3. This problem calls for the binomial distribution, $X~\sim Bin(n=5, p)$, however, we do not know $p$. This is the probability that a single piece is produced in 4 minutes or less. $p=P(X<4)$ with $X\sim Exp(0.2)$. In this case $p=P(X<4)=1-e^{-0.2\times 4}\approx 0.55$. Then, binomial $X\sim Bin(5, 0.55)$, and then $P(X=2)\approx 0.2757$.

:::

[Back to the exercise](#exp-example-1)

## Exponential Exercise {#exp-example-sol-2}

In this case the r.v. $X$ is the time until the first consultation, and $X\sim Exp(0.1)$.

$$P(X>10)=1-P(X<10)=1-F(10)$$

Or

$$1-(1-exp^{-0.1\times 10})=e^{-1}\approx 0.3679$$


[Back to the exercise](#exp-example-2)

