---
title: "Chapter II Part I - Statistics I - Exercises"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, include = FALSE}
library(lubridate)
y <- year(Sys.Date())-1
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }
</style>

<!-- Style Block for nice tables -->


This is a set of exercises created by the teaching Faculty for Statistics I, Chapter II - Introduction to Probability Theory, from the Lisbon Accounting and Business School.

## Questions

### Question 1

::: {.webex-check .webex-box}

At a Christmas market, a certain stall sells small souvenirs. Let $X$ a *r.v.* that counts the number of souvenirs sold in that stall each hour. Let the *pdf* for $X$ be:

:::{.table}
| $x$ | 0 | 1 | 2 | 3 | 4 |
| :--: | :--: | :--: | :--: | :--: | :--: |
|$f(x)$ | 0.1 | 0.2 | 0.3 | 0.2 | 0.2 |
:::

1. `r torf(TRUE)` The probability of selling, within the hour, 3 souvenirs, given that at least 1 was sold is $0.(2)$.
2. `r torf(FALSE)` The events "selling an odd number of souvenirs" and "selling more than 2 souvenirs" are independent.

:::

### Question 2

::: {.webex-check .webex-box}

Consider the following random variable $X$ with its $pdf$ given by:

::: {.table}
| $x$ | 0 | 1 | 2 | 3 | 4 | 
| ---  | :---:   | :---:   | :---:   | :---:   | :---:   |
|$f(x)$ | 0.05 | $\alpha$ | 0.35 | 0.25 | 0.05 |
:::

1. `r torf(TRUE)` $\alpha$ must equal 0.3.

2. `r torf(FALSE)` It must bet the case that $P(X=2|X\leq 3)=0.632$.

3. `r torf(TRUE)` By contrasting the mean and the median of $X$ we can conclude that the distribution is asymmetric.

:::

### Question 3

::: {.webex-check .webex-box}

Let $X$ be a *r.v.* with $cdf$ given by:
$$
F(X)=\begin{cases}
0 & x\leq -1 \\
0.25 & -1\leq x < 0 \\
a & 0 \leq x < 1 \\
0.6 & 1\leq x < 2 \\
0.85 & 2 \leq x < 3 \\
1 & \geq 3
\end{cases}
$$


1. Knowing that $P(X=0)=0.15$, $a$ must be equal to `r mcq(c(answer="0.40","0.35","0.45","None"))`

2. `r torf(FALSE)` $P(X>1|0\leq X\leq 3)=0.375$

3. `r torf(FALSE)` By looking to the relative dispersion we can conclude that $E[X]$ is representative of the balance of the distribution.

:::

### Question 4

::: {.webex-check .webex-box}

The number of work absences, in days, of the workers of a company is a random variable with the following *pdf*

::: {.table}
| $x$ | 1 | 2 | 3 | 4 | 5  | $x\notin\{1,2,3,4,5\}$ |
| ---  | :---:   | :---:   | :---:   | :---:   | :---:   | :---:  | 
|$f(x)$ | **a** | 0.15 | 0.3 | 0.10 | **b** | 0 |
:::

It is known further that 60% of the workers did not skip more than 3 days of work.

1. The values for **a** and **b** are `r mcq(c(answer="a=0.15, b=0.3", "a=0.3, b=0.15", "a=0.1, b=0.35", "None of the above"))`

2. `r torf(TRUE)` By looking at the distribution function for $X$, we can find that the probability that a worker skipped 5 days of work, given that it skipped more than 3 days is equal to 0.75.

3. After a strike at the company, all workers skipped 1 day of work. The impact on the expected value for skipped days is `r mcq(c("Null","decreasing",answer="increasing", "we cannot say anything"))`

:::

### Question 5

::: {.webex-check .webex-box}

Let $X$ a *r.v.* with the following *cdf*:

$$
F(x)=\begin{cases}
0 & x < 1 \\
1/6 & 1\leq x < 2 \\
1/4 & 2 \leq x < 4 \\
1/2 & 4 \leq x < 5 \\
7/12 & 5 \leq x < 6 \\
1 & x \geq 6
\end{cases}
$$

1. The first and third quartile are: `r mcq(c("1 and 5",answer="2 and 6","2 and 5","Non of the above"))`
2. `r torf(TRUE)` $P(X=2)=1/12$
3. `r torf(FALSE)` $F(3)=0$ Because $3$ is not a value in the support for $X$.

:::

### Question 6

::: {.webex-check .webex-box}

Let $X$ a *r.v.* with the following *pdf*:

$$
f(x)=\begin{cases}
\theta x^2 & 0\leq x < 1 \\
0 & \mathbb{R}\setminus[0,1]
\end{cases}
$$

1. $\theta$ is equal to `r mcq(c("1/3",answer="3","1","None"))`
2. `r torf(TRUE)` The *cdf* is given by: $$ F(x)=\begin{cases} 0 & x < 0 \\ x^3 & 0\leq x < 1 \\ 1 & x\geq 1\end{cases} $$
3. The median for $X$ is `r mcq(c("0.5","0.2063",answer="0.7937","None of the above"))`
3. `r torf(FALSE)` The mean and standard deviation of $X$ are $E[X]=0.75$ and $V[X]=0.1936$

:::

### Question 7

::: {.webex-check .webex-box}

Given the *cdf* for the r.v. $X$: 
$$
F(x)=\begin{cases}
0 & x <0\\
x/3 & 0\leq x < 3\\
1 & x\geq 3
\end{cases}
$$

1. `r torf(TRUE)` $P(1\leq X \leq 2)=1/3$
2. `r torf(FALSE)` The mean for $X$ is 2.5

:::

### Question 8

::: {.webex-check .webex-box}

Let $X$ a *r.v.* with *cdf*:

$$
F(x)=\begin{cases}
0 & x <1\\
x-1 & 1\leq x < 2\\
1 & x\geq 2
\end{cases}
$$

1. `r torf(TRUE)` It is a continuous *r.v.*
2. `r torf(TRUE)` The median is 1.5
3. `r torf(FALSE)` $P(X\leq 1.6 | X > 1.2)=0.4$

:::

### Question 9

::: {.webex-check .webex-box}

Let $X$ a *r.v.* absolutely continuous with the following *pdf*:

$$
F(x)=\begin{cases}
0 & x<8\\
\frac{(x-8)^2}{2} & 8\leq x < 9 \\
1-\frac{(10-x)^2}{2} & 9\leq x < 10 \\
1 &  x\geq 10
\end{cases}
$$

1. $P(X\geq 9.2 | 8.5 \leq X \leq 9.5)$ is equal to `r mcq(c("0.195","0.875",answer="0.26","None of the above"))`

:::

### Question 10

::: {.webex-check .webex-box}

At SuperStore :convenience_store:, three trained employees are qualified to operate the checkout counters, restock products on the shelves, and perform some administrative tasks. SuperStore has three checkout counters, and at least one of them must always be operating.

At any given day and moment when SuperStore is open to customers, consider the following random variables:

* $X$ N of employees in the checkout counters :shopping_cart: :credit_card: .
* $Y$ N of employees restocking products on the shelves :package:.

The r.v. $X$ has $\Omega_X=\{1,2,3\}$

1. If $F_X$ is the *cdf* for $X$, then $P(2\leq X\leq 2.5)$ might be computed through: `r mcq(c("a", answer="b","c","d"))`
    a) $F_X(2.5)-F_X(2)$
    b) $F_X(2.5)-F_X(2^-)$
    c) $F_X(2.5^-)-1+F_x(2)$
    d) $F_X(2)-F_X(2.5)$

For the next questions consider:

:::{.table}
| $x$ | 1 | 2 | 3 |
| :--: | :--: | :--: | :--: |
| $f_X(x)$ | 0.17 | 0.8 | 0.03 |
:::

2. The distribution function for $X$ at 2.1 has the value: `r mcq(c("0.8", answer="0.97","0.17","None"))`

3. $E[10X-0.6]$ equals `r mcq(c(answer="18","19.4","18.6","None"))`
4. $E[1/x]$ equals `r mcq(c(answer="0.58","108.38","0.538","None"))`

Consider the following table for the joint probability of $(X,Y)$

::: {.table}
|$X\setminus Y$ | 0 | 1 | 2 | |
| :--: | :--: | :--: | :--: | :--: |
| 1 | $a$ | $2b$ | $b$ | |
| 2 | 0.1  | $c$ | 0 | |
| 3 | 0.03 | 0 | 0 | |
|   |   |  |  | 1| 
:::

5. If $P(X=1|Y=0)=0.05$ then `r mcq(c("b=0.06 and c=0.7","b=0.08 and c=0.4",answer="b=0.05 and c=0.7","None"))`

For the following lines consider $a=0.02$ and $b=0.05$ and $c=0.7$

6. $P(X=2|Y\geq 1)$ is approximately `r mcq(c(answer="0.824","0.941","0.875","None"))`
7. `r torf(FALSE)` $X$ and $Y$ are independent *r.v.s.*
8. Knowing that $E[Y]=0.9$, $cov(X,Y)$ equals `r mcq(c("-14.6",answer="-0.074","0.074","None"))`

:::

### Question 11

::: {.webex-check .webex-box}

Consider the following information about the random pair $(X,Y)$:

- $X$ has support $\Omega_X=\{0,1\}$ and $Y$ has support $\Omega_Y=\{-1,1\}$
- $P(X=0)=0.5$ and $P(Y=1)=0.6$
- $P(X=1,Y=1)=p$ with $0<p<1$

1. $p$ must equal `r mcq(c("0.2",answer="0.3","0.4","None"))` to make $E[XY]=0.1$
2. If $p=0.4$, then $P(X+Y=0)$ equals `r mcq(c("0.4","0.5",answer="0.1","None"))`
3. If $p=0.4$ $V[X]=0.25$ and $Cov(X,Y)=0.2$, $V[2Y-X]$ equals `r mcq(c("1.77","3.19",answer="3.29","None"))`

:::

### Question 12

::: {.webex-check .webex-box}

Consider the *r.v.* $X$ associated with the distribution function $F$

1. `r torf(FALSE)` It must be that $P(X>k)+P(X<k)=1$ $\forall k$
2. `r torf(TRUE)` If $a<b$ then $P(a\leq X\leq b)=F(b)-F(a)+P(X=a)$
3. `r torf(TRUE)` If $X$ is discrete, with *pdf* $f$, then $F(a)\geq f(a)$ $\forall a$.

:::

### Question 13

::: {.webex-check .webex-box}

Consider the random pair $(X,Y)$ and let $a,b$ be two scalars ($\in\mathbb{R}$)

1. `r torf(TRUE)` If $V[X]$ exists, then $V(X-a)=V(a-X)$
2. `r torf(TRUE)` If $P(Y=aX+b)=1$ then $|\rho|=1$
3. `r torf(FALSE)` If $X$ and $Y$ are discrete, then it is always true that $E[XY]=E[X]E[Y]$

:::
