---
title: "Chapter III Part II - Statistics I - Exercises"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, include = FALSE}
library(lubridate)
y <- year(Sys.Date())-1
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }
</style>

<!-- Style Block for nice tables -->


This is a set of exercises created by the teaching Faculty for Statistics I, Chapter III, part II, continuous probability models, from the Lisbon Accounting and Business School.


## Questions

### Question 1

::: {.webex-check .webex-box}

The length of short commercial spots (between 5 and 12 sec) on a streaming platform can be considered a random variable with uniform distribution.

1. `r torf(TRUE)` The distribution function is given by $$F(x)=\begin{cases}0& x\leq 5\\ \frac{x-5}{7}& 5<x\leq 12\\ 1&x\geq 12\end{cases}$$
2. The probability that one of these spots last at least 7 seconds is `r mcq(c("2/7",answer="5/7","6/7","None"))`
3. The probability that one of these spots last more than 6 seconds, given that it will not last more than 10 seconds is `r mcq(c(answer="4/5","4/7","1/7","None"))`
4. `r torf(TRUE)` The expected value and standard deviation for this r.v. is $8.5$ and $2$ respectively, both in seconds.

:::

### Question 2

::: {.webex-check .webex-box}

Let the r.v. $X\sim U(2,b)$ with $b>2$.

1. $b$ must equal what in order for $P(3\leq X\leq 5)=0.4$? `r mcq(c("12","8",answer="7","None"))`

:::

### Question 3

::: {.webex-check .webex-box}

In a factory, the time to produce some item is a r.v. distributed exponentially with expected value of 5 minutes.

1. $\lambda$ in this case is equal to `r mcq(c("5",answer="1/5","1/25","None"))`
2. You now that the item has been in production for at least 2 minutes. The probability of taking at least 4 more minutes until completed is `r mcq(c("0.5507","0.6703",answer="0.4493","None"))`
3. `r torf(TRUE)` Choose randomly 5 items. The probability that two of them took less than 4 minutes to produce is 0.2757.

:::

### Question 4

::: {.webex-check .webex-box}

Let $X$ be a r.v. distributed exponentially with $\lambda=0.5$.

1. `r torf(TRUE)` $P(X>2)=e^{-1}$

:::

### Question 5

::: {.webex-check .webex-box}

On the consultation of Dr. Rotcod The time until the first appointment, and between consecutive appointments, are independent random variables distributed exponentially with parameter equal to 0.1.

1. `r torf(TRUE)` The probability of starting the first appointment after the first 10 minutes is 0.3679.

:::

### Question 6

::: {.webex-check .webex-box}

The number of cars checked in the inspection center, in an hour, is a r.v. distributed Poisson with an expected value of 4.

1. `r torf(FALSE)` The expected time that a car will be waiting until being inspected is 25 minutes.

:::

### Question 7

::: {.webex-check .webex-box}

Let $Z\sim\mathcal{N}(0,1)$

1. `r torf(FALSE)` $P(0<Z\leq 2.05)=0.9790$
2. `r torf(TRUE)` $P(-1.22\leq Z < 1.05)=0.7419$
3. `r torf(FALSE)` $P(Z\geq -2.05)=0.0202$
4. `r torf(TRUE)` The value for $k$ such that $P(|Z|>k)=0.05$ is equal to 1.96.
5. `r torf(TRUE)` The value for $k$ such that $P(|Z|<k)=0.90$ is equal to 1.645.

:::

### Question 8

::: {.webex-check .webex-box}

If $X\sim \mathcal{N}(\mu=6,\sigma^2=25)$ then 

1. `r torf(FALSE)` $P(6<X\leq 12)=0.8849$
2. `r torf(TRUE)` $P(0\leq X \leq 8) = 0.5404$
3. `r torf(TRUE)` $P(X\leq -4)=0.0228$
4. `r torf(TRUE)` $P(|X-6|>10)=0.0456$
5. `r torf(FALSE)` $k$ must be equal to 0.4 such that $P(X>k)=0.90$.

:::

### Question 9

::: {.webex-check .webex-box}

Let $X$ a r.v. distributed $\mathcal{N}(12,2)$. Considere $P(a\leq X\leq 15)=0.7332$

1. `r torf(TRUE)` $a$ must be equal to 10.32.

:::

### Question 10

::: {.webex-check .webex-box}

The number of deposits made in a fintech, every day, is a r.v. distributed $\mathcal{N}(120,64)$ in euro.

1. `r torf(FALSE)` The percentage of days in which the amount of the deposits is in between 105 and 135 euro is 96.99%
2. `r torf(TRUE)` The probability that the amount of deposits is above its mean, in the days that it never surpasses 125 is 0.32.
3. `r torf(TRUE)` The mean and variance of the weekly total amount of deposits (i.e. in 5 days) are respectively 600 euro and 320 euro<sup>2</sup>.

:::

### Question 11

::: {.webex-check .webex-box}

An item is produce in a way such that its weight (in grams) is a r.v. $X\sim \mathcal{N}(250,100)$.

1. The probability of a package weight between 230.4 and 269.6 grams is `r fitb(0.95)` (use 2 decimals, like 0.12)
2. The quality control department rejects packages with a weight farther away from the mean than 15.3 grams. The probability that a random package is rejected is `r fitb(0.126)` (use three decimals, like 0.123)
3. Find and interpret the value for $k$ such that $P(X<k)=0.9$ Answer: `r fitb(262.82)` grs (use only 2 decimals, like 123.45)
4. Randomly picking 4 packages, compute the probability that the total weight is above its mean. Justify. Answer: `r fitb(0.5)`

:::

### Question 12

::: {.webex-check .webex-box}

Being $X$ and $Z$ independent random variables, distributed $X\sim\mathcal{N}(1,1)$ and $Z\sim\mathcal{N}(0,1)$

1. `r torf(TRUE)` $P(X<Z+1)=0.5$ 
2. The value for $a$ such that $P(X>Z+a)=0.1$ is equal to `r mcq(c("1.282",answer="2.813","1","None"))`

:::


### Question 13

::: {.webex-check .webex-box}

Check the truthfulness of each sentence

1. `r torf(TRUE)` If $Y\sim U(0,1)$ and $0<x<z<1$, then: $$P(x<Y<z)=z-x$$
2. `r torf(TRUE)` If $X\sim U(0,1)$ with $0<a<b<1$, then $$P(X<a|X<b)=\frac{a}{b}$$
3. `r torf(TRUE)` If $X$ is normally distributed, then the media coincides with the mean. 
4. `r torf(TRUE)` If $X\sim \mathcal{N}(0,1)$, and $Y\sim\mathcal{N}(1,2)$, with $X$ and $Y$ independent r.v.s, then $2X+Y\sim\mathcal{N}(1,6)$
5. `r torf(FALSE)` Letting $\Phi$ be the distribution function of a r.v. distributed normal standard, then $\Phi(-x)=\Phi(x)$ $\forall x\in\mathbb{R}$.
6. `r torf(FALSE)` Letting $\phi$ represent the density of a r.v. distributed normal standard, then $\phi(-x)=1-\phi(x)$ $\forall x\in\mathbb{R}$.

:::











