---
title: "Chapter II Part I - Statistics I - Exercises"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, include = FALSE}
library(lubridate)
y <- year(Sys.Date())-1
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }
</style>

<!-- Style Block for nice tables -->


This is a set of exercises created by the teaching Faculty for Statistics I, Chapter II - Random Variables, from the Lisbon Accounting and Business School.

## Questions

### Question 1

Consider the following events:

A.  "get a 7 when rolling a cubic die"
B.  "Spain win the next World Cup"
C.  "rain in London"

```{r}
q1 <- c(
  'A is impossible, while C is unlikely',
  'B is certain, and C is very likely',
  answer='$P(B)$ and $P(C)$ are computed using the subjective notion of probability',
  'A and C are impossible'
)
```

The correct choice is: `r mcq(q1)`

### Question 2

```{r}
q2 <- c(
  'the frequentist definition of probability',
  answer = 'the classic definition of probability',
  'the subjective definition of probability',
  'the axiomatic definition of probability'
)

```

When computing a probability from the analysis of all possible outcomes (equally likely) it was used: `r mcq(q2)`

### Question 3

```{r}
q3 <- c(
  answer = '0',
  'P(A)',
  'P(B)',
  '1'
)
```

Let $\Omega$ be the outcome space associated to a random experiment. Let $A$ and $B$ two events ($A\subset\Omega$ and $B\subset \Omega$) such that $0<P(A)<1$ and $0<P(B)<1$. You know that $A\subset B$. What is $P[(A\cup B)\cap B^c]$?

`r mcq(q3)`

### Question 4

Let $A,B,C$ events defined in $\Omega$ such that:

1. $A\cup B\cup C=\Omega$
2. $P(A)=0.3$
3. $P(C)=0.5$
4. $P(B^c)=0.7$
5. $A\cap B=\emptyset$
6. $B\cap C=\emptyset$

`r torf(TRUE)` $P(A\cap C)=0.1$

### Question 5

Let $A$ and $B$ two independent events defined in the same $\Omega$, where $A$ has a probability twice as large as $B$. Knowing that $0.5$ is the probability that at least $A$ or $B$ happens, you have:
a) $P(B)=\frac{3+\sqrt{5}}{4}$
b) $P(B)=\frac{3-\sqrt{5}}{4}$
c) $P(B)=\frac{3+\sqrt{5}}{4}$ or $P(B)=\frac{3-\sqrt{5}}{4}$
d) None of the above

```{r}
q5 <- c(
  'a',
  answer = 'b',
  'c',
  'd'
)
```

`r mcq(q5)`

### Question 6

Let $A$ and $B$ be two events in the same outcomes space $\Omega$ such that $P(A)=0.75$, $P(B)=0.5$ and $P(A\cup B)=1$. Then, $P(A|B)$ is equal to:

```{r}
q6 <- c(
  '13',
  '34',
  answer = '12',
  '14'
)
```
`r mcq(q6)`

### Question 7

Let $A$ and $B$ two events in the same sample space $\Omega$. You know that $P(A)=a$ and $P(B)=b$, with $0<a<1$ and $0<b<1$. The probability that neither $A$ nor $B$ happen is:

a) `r torf(TRUE)` $(1-a-b+ab)$ if $A$ and $B$ are independent events.
b) `r torf(FALSE)` $1$ if $A$ and $B$ are complementary events.

### Question 8

### Question 9

### Question 10

