---
title: "Suggested exercises for correspondences"
format: 
  html:
    css: [Exercises/webex.css]
    include-after-body: [Exercises/webex.js]
    embed-resources: true
lang: en
editor: source
page-layout: full
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, include = FALSE}
library(lubridate)
y <- year(Sys.Date())-1
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

<!-- Style Block for nice tables -->

<style>
  table {
    width: 100%;
    border-collapse: collapse;
    text-align: center;
    font-family: Arial, sans-serif;
  }

  thead tr:first-child th {
    background-color: #f2f2f2;
    border-bottom: 2px solid #999;
  }

  thead tr:nth-child(2) th {
    background-color: #e0e0e0;
    border-bottom: 2px solid #999;
  }

  tbody td {
    padding: 8px;
    border-bottom: 1px solid #ccc;
  }

  tbody tr:nth-child(even) td {
    background-color: #f9f9f9;
  }

  th, td {
    vertical-align: middle;
  }
</style>

<!-- Style Block for nice tables -->

### Question 1

1. Analyze if the following correspondences are uhc and lhc.

    1. $\Gamma:[0,1]\twoheadrightarrow [0,1]$, $\Gamma(x)=(a,b)\subset [0,1]$.
    2. $\Gamma:[0,1]\twoheadrightarrow [0,1]$, $\Gamma(x)=(0.2+0.2x,0.4+0.2x)\subset [0,1]$.
    3. $\Gamma:[0,1]\twoheadrightarrow [0,1]$, $\Gamma(x)=[0.2+0.2x,0.4+0.2x]\subset [0,1]$.

2. Propose a correspondence that:

    1. It is neither uhc nor lhc.
    2. It is both, *i.e.* it is continuous.